---
title: "Examining background gene lists using simulation - comparison of different enrichment techniques"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

```{r,libs}

library("parallel")
library("edgeR")
library("DESeq2")
library("limma")
library("stringi")
library("kableExtra")
library("fgsea")
library("clusterProfiler")
source("simpw_func.R")

```

## Intro

TODO


xxx object slots

1. expression counts

2. ground truth up genes

3. ground truth down genes

4. ground truth up gene sets

5. ground truth down gene sets

6. DE result (DESeq2)

7. DE genes up observed

8. DE genes down observed

9. clusterprofiler_default up gene sets

10. clusterprofiler_default down gene sets

11. clusterprofiler_fixed up gene sets

12. clusterprofiler_fixed down gene sets

13. clusterprofiler_nobg up gene sets

14. clusterprofiler_nobg down gene sets

15. fora up gene sets

16. fora down gene sets

17. fgsea up gene sets

18. fgsea down gene sets

19. mitch up

20. mitch dn

# Get count data

```{r,getcounts,warning=FALSE}

a <- countData()

```

# Generate gene sets

```{r,genesetgeneration}

gsets <- randomGeneSets(a,setsize=30,nsets=200)

```

# run simulations over a range of parameters

```{r,setparameters}

SIMS=1000
#run 500 sims
FRAC_DE=0.05
FC=1
N_REPS=3
DGE_FUNC="deseq2"
SUM_COUNT=2e7
VARIANCE=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1)

mygrid <- expand.grid(FRAC_DE,FC,N_REPS,DGE_FUNC,SUM_COUNT,VARIANCE)
colnames(mygrid) <- c("FRAC_DE","FC","N_REPS","DGE_FUNC","SUM_COUNT","VARIANCE")

mygrid

```

Now run the analysis.

```{r,sims}

res <- lapply(1:nrow(mygrid), function(i) {
  FRAC_DE=mygrid[i,"FRAC_DE"]
  FC=mygrid[i,"FC"]
  N_REPS=mygrid[i,"N_REPS"]
  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
  SUM_COUNT=mygrid[i,"SUM_COUNT"]
  VARIANCE=mygrid[i,"VARIANCE"]
  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets,cores=32)
  as.data.frame(do.call(rbind, x))
})

res <- do.call(rbind,res)

saveRDS(res,file="res.Rds")

```

Now show the results.

```{r,results}

res %>% kbl(caption="simulation_results") %>% kable_paper("hover", full_width = F)

```

## Plotfig1_ORAjac.png

```{r,plot1}

cp <- subset(res,PWAY_FUNC == "clusterprofiler")
fo <- subset(res,PWAY_FUNC == "fora")
fg <- subset(res,PWAY_FUNC == "fgsea")

cp %>% kbl(caption="clusterprofiler") %>% kable_paper("hover", full_width = F)
fo %>% kbl(caption="fora") %>% kable_paper("hover", full_width = F)
fg %>% kbl(caption="fg") %>% kable_paper("hover", full_width = F)

par(mfrow=c(1,3))
par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")

legend("bottomleft", inset=.02, title="tool",
   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
   horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")

plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")


png("fig3_sim.png",width=720, height=480)
par(mar=c(c(5.1, 5.1, 2.1, 2.1) ))
par(mfrow=c(1,3))

plot(cp$VARIANCE,cp$p,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="precision")
points(fo$VARIANCE,fo$p,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$p,type="b",pch=19,col="blue")
grid()
legend("bottomleft", inset=.02, title="tool",
   c("clusterprofiler","fora","fgsea"), col=c("black","red","blue"),
   horiz=FALSE, cex=1.2, pch=19,lwd=2)

plot(cp$VARIANCE,cp$r,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="recall")
points(fo$VARIANCE,fo$r,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$r,type="b",pch=19,col="blue")
grid()
plot(cp$VARIANCE,cp$f,ylim=c(0,1),type="b",pch=19,xlab="variance added",ylab="index",main="f1")
points(fo$VARIANCE,fo$f,type="b",pch=19,col="red")
points(fg$VARIANCE,fg$f,type="b",pch=19,col="blue")
grid()
dev.off()

```

## Session information


```{r,session}

sessionInfo()

```
