---
title: "Examining background gene lists using simulation - comparison of different enrichment techniques"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

```{r,libs}

library("tidyverse")
library("parallel")
library("edgeR")
library("DESeq2")
library("limma")
library("stringi")
library("kableExtra")
source("simpw_func.R")

```

# Intro

TODO


xxx object slots

1. expression counts

2. ground truth up genes

3. ground truth down genes

4. ground truth up gene sets

5. ground truth down gene sets

6. DE result (DESeq2)

7. DE genes up observed

8. DE genes down observed

9. fgsea up gene sets

10. fgsea down gene sets

11. phyper up gene sets

12. phyper down gene sets

13. clusterprofiler_default up gene sets

14. clusterprofiler_default down gene sets

15. clusterprofiler_fixed up gene sets

16. clusterprofiler_fixed down gene sets

17. clusterprofiler_nobg up gene sets

18. clusterprofiler_nobg down gene sets

19. fisher up gene sets

20. fisher down gene sets

21. DOSE up gene sets

22. DOSE down gene sets

# Get count data

```{r,getcounts,warning=FALSE}

a<-countData()

```

# Generate gene sets

```{r,genesetgeneration}

gsets<-randomGeneSets(a,setsize=50,nsets=500)

```

# run simulations over a range of parameters

```{r,setparameters}

SIMS=8
FRAC_DE=0.05
FC=1
N_REPS=4
DGE_FUNC="deseq2"
SUM_COUNT=3e7
VARIANCE=c(0.25,0.3,0.35,0.4,0.45)

mygrid <- expand.grid(FRAC_DE,FC,N_REPS,DGE_FUNC,SUM_COUNT,VARIANCE)
colnames(mygrid) <- c("FRAC_DE","FC","N_REPS","DGE_FUNC","SUM_COUNT","VARIANCE")

mygrid

```

Now run the analysis.

```{r,sims}

res <- lapply(1:nrow(mygrid), function(i) {
  FRAC_DE=mygrid[i,"FRAC_DE"]
  FC=mygrid[i,"FC"]
  N_REPS=mygrid[i,"N_REPS"]
  DGE_FUNC=as.character(mygrid[i,"DGE_FUNC"])
  SUM_COUNT=mygrid[i,"SUM_COUNT"]
  VARIANCE=mygrid[i,"VARIANCE"]
  x <- agg_dge(a,N_REPS,SUM_COUNT,VARIANCE,FRAC_DE,FC,SIMS,DGE_FUNC,gsets)
  as.data.frame(do.call(rbind, x))
})

res <- do.call(rbind,res)

```

Now show the results.

```{r,results}

res %>% kbl(caption="simulation_results") %>% kable_paper("hover", full_width = F)

```

# Session information


```{r,session}

sessionInfo()

```
